---
title: "Data Visulization in R"
author: "Yiming Zhang"
date: "`r Sys.Date()`"
output: ioslides_presentation
widescreen: true
smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 7, 
                      fig.height = 3.5, 
                      fig.align = "center", warning = FALSE)

package.required <- c("ggplot2", "ggrepel", "GGally", "ggvis", "ggExtra", "gganimate", "gridExtra")

for(i in 1:length(package.required)){
  if(!package.required[i] %in% rownames(installed.packages())){
    install.packages(package.required[i])
  }
}

## Load packages
library(ggplot2)
library(ggrepel)
library(ggExtra)
library(GGally)
library(gganimate)
library(ggvis)
library(gridExtra)
library(gifski)


## Read data
nba.data <- read.csv("NBA_Team_Data.csv")
nba.data <- nba.data[, c(2, 3, 7, 9, 30, 31, 42, 44, 45)]
mean.win <- aggregate(WIN. ~ SEASON + REGION, FUN = mean, data = nba.data)



practice1 <- ggplot(data = nba.data, aes(x = OFFRTG, y = DEFRTG, size = WIN.)) +
  geom_point(aes(color = REGION), shape = 1) + 
  geom_text(data = nba.data[which(nba.data$WIN. >= 0.6), ], aes(label = ABV), size = 1.5) +
  geom_text(data = nba.data[which(nba.data$WIN. <= 0.3), ], aes(label = ABV), size = 1.5) +
  facet_wrap(~SEASON) +
  theme_bw() +
  scale_y_reverse(limits = c(120, 97)) +
  scale_color_manual(values = c("blue3", "red3")) +
  theme()

practice2 <- ggplot(data = nba.data, aes(x = REGION, y = WIN., color = REGION)) + 
  geom_violin(fill = NA) + 
  geom_boxplot(fill = NA, width = 0.35, outlier.shape = 1) +
  geom_point(data = mean.win, aes(x = REGION, y = WIN.), shape = 5) +
  facet_wrap(~SEASON, nrow = 1) +
  scale_color_manual(values = c("blue3", "red3")) +
  theme_bw()
```


## Outline

- General introduction to data visualization

- Introduction to ggplot2 in R

- Practices with ggplot2

- Introduction to the extension of ggplot2

# General introduction to data visualization

## When to visualize data

- Exploratory data analysis

  + Explore pattern, trend, and distribution
  
  + Find correlation between variables
  
  + Regression analysis

- Report statistical results

- Communicate with non-statisticians

  + Share findings

  + Show fancy plots to your audiences

## What to plot

- One variable: Histogram, Bar chart, Density plot...

- Two variables: Scatter plot, Box plot, Violin Plot...

- Multiple variables: Heatmap...

- Think of your data and variable carefully, and choose the most appropriate statistical plot.

## Why we need data visualization

- Better summary of statistics than table and text.

- Easy to show a trend or a pattern in the data.

- A more interesting way to catch your audiences' eyes.

- For fun...

## Introduction to the data set {.smaller}

- All data are from http://stats.nba.com

```{r}
head(nba.data)
```

- WIN.: Winning rate, which is the percentage of games played that a team has won.

- PTS: The number of points scored.

- OFFRTG: Offensive Rating, which measures a team's points scored per 100 possessions.

- DEFRTG: Defensive Rating, which is the number of points allowed per 100 possessions by a team.

- PACE: Pace, which is the number of possessions per 48 minutes for a team.

- REGION: East/West.

- ABV: The abbreviation of a team.

## What information can we get from this plot?

```{r,  fig.width=8, fig.height = 5}
practice1
```

## What information can we get from this plot?

```{r,  fig.width=8, fig.height = 5}
practice2
```

## But "bad" plots may....

- be hard to read if labels and legends are not clear

- confuse people if it is not well-designed

- deliver misleading information (sometimes in purpose)

## What's the issue in this plot?

```{r,  fig.width=8, fig.height = 5}
ggplot(data = mean.win, aes(x = SEASON, y = WIN., fill = REGION)) + 
  geom_bar(position = "dodge", stat = "identity", width = 0.5) +
  scale_fill_discrete("", labels = NULL)
```

## Add labels for the legend and the y axis

```{r,  fig.width=8, fig.height = 5}
ggplot(data = mean.win, aes(x = SEASON, y = WIN., fill = REGION)) + 
  geom_bar(position = "dodge", stat = "identity", width = 0.5) +
  scale_y_continuous("Mean Winning Rate")
```

## What's the issue in this plot?

```{r,  fig.width=8, fig.height = 5}
ggplot(data = nba.data, aes(x = OFFRTG, y = DEFRTG)) + 
  geom_point() +
  geom_smooth(method = "lm") + 
  facet_wrap(REGION~SEASON, nrow = 2, scales = "free")
```

## Compare in the same scale

```{r,  fig.width=8, fig.height = 5}
ggplot(data = nba.data, aes(x = OFFRTG, y = DEFRTG)) + 
  geom_point() +
  geom_smooth(method = "lm") + 
  facet_grid(REGION~SEASON)
```

## How to visualize data in R?

- `graphics`: The R basic graphics package

- `ggplot2`: The grammar of graphics

- `plotly`: Interactive plot in RShiny

- `leaflet`: Interactive maps

## Basic graphical package in R

## The `plot` function 

# Introduction to ggplot2

## ggplot2

- Grammar of graphic

- Both quick and complex plot in an easy way

- Nice aesthetic settings

- Great docummentation and tons of online instructions

## An example of ggplot2 {.smaller}

The histogram of winning rate in different regular NBA seasons and regions:

```{r, echo=TRUE}
ggplot(data = nba.data, aes(x = WIN.)) + 
  geom_histogram(binwidth = 0.1, color = "black") + facet_grid(REGION ~ SEASON)
```

## Comparing with `graphics` package

```{r, fig.width = 8, fig.height = 5}
par(mfrow = c(2, 4), mar = c(2, 2, 3, 1))
for(i in levels(nba.data$REGION)){
  for(j in levels(nba.data$SEASON)){
    subdata <- subset(nba.data, REGION == i & SEASON == j)
    hist(subdata$WIN., breaks = seq(0.05, 0.95, 0.1),
         xlim = c(0, 1),
         main = paste(i, j, sep = " ,"))
  }
}
```

## Code for the same plot

```{r, echo = TRUE, eval = FALSE}
par(mfrow = c(2, 4), mar = c(2, 2, 3, 1))
for(i in levels(nba.data$REGION)){
  for(j in levels(nba.data$SEASON)){
    subdata <- subset(nba.data, REGION == i & SEASON == j)
    hist(subdata$WIN., breaks = seq(0, 1, 0.1),
         main = paste(i, j, sep = " ,"))
  }
}
```

## Grammar of Graphics {.smaller}

- Idea: graph is a combination of independent building blocks.

- **Data** that you want to visualise and a set of **aesthetic** mappings describing how variables in the data are mapped to aesthetic attributes.

- **Layers** made up of geometric elements and statistical transformation. Geometric objects, **geoms** for short, such as points, lines, polygons, etc. Statistical transformations, **stats** for short, summarise data in many useful ways.

- The **scales** map values in the data space to values in an aesthetic space, whether it be colour, or size, or shape.

- A **coordinate system**, coord for short, describes how data coordinates are mapped to the plane of the graphic.

- A **facet** describes how to break up the data into subsets and how to display those subsets as small multiples. 

- A **theme** which controls the finer points of display, like the font size and background colour. 

## The strat of plotting a graph {.smaller}

- `ggplot()` is always the first line of the ggplot.

- We can specify the **data** set and the **aesthetics** mapping variables in the `ggplot()`.

```{r, echo=TRUE}
p <- ggplot(data = nba.data, aes(x = OFFRTG, y = WIN.))
p
```

## Aesthetics

- Map the variables in the data to the components in the plot

- x: x axis

- y: y axis

- color: color of the boundary of a figure

- fill: color of the inside of a figure

- shape: shape of points, solid point, circle, triangle...

- size: size of points

- linetype: type of lines, solid line, dashed line...

- ...

## Geometries {.smaller}

- Geometries are the actual graphical elements displayed in a plot. They can visualize the mapping variables (specified in `aes()`) from the data.

- We use `+` to connect multiple geometrics

```{r, echo = TRUE}
p + geom_point()
```

## Geometries

- We can also specify `data` and `aes` in `geom` function. They don't have to be the same as those in `ggplot()`.

```{r, echo = TRUE}
ggplot() + geom_point(data = nba.data, aes(x = DEFRTG, y = WIN.))
```

## `geom` function {.smaller}

- One variable

```{r, fig.show = "hold", out.width = "50%", fig.align='default', echo=TRUE}
p <- ggplot(data = nba.data, aes(x = WIN.))
p + geom_histogram(binwidth = 0.1)
p + geom_density()
```


## `geom` function {.smaller}

- Continuous X, continuous Y

```{r, fig.show = "hold", out.width = "35%", fig.align='default', echo=TRUE, fig.height=3.5, fig.width=7}
p <- ggplot(data = nba.data, aes(x = OFFRTG, y = WIN.))
p + geom_point(); p + geom_line(); p + geom_density_2d(); p + geom_smooth(method = "lm")
```

## `geom` function {.smaller}

- Discrete X, continuous Y

```{r, fig.show = "hold", out.width = "50%", fig.align='default', echo=TRUE}
p <- ggplot(data = nba.data, aes(x = SEASON, y = WIN.))
p + geom_boxplot()
p + geom_violin()
```

## `geom` function {.smaller}

- Plot text on the graph: aes(x, y, label)

```{r, echo=TRUE, fig.width=8, fig.height=4}
p <- ggplot(data = nba.data, aes(x = OFFRTG, y = WIN.)) + 
  facet_wrap(~SEASON)
p + geom_text(aes(label = ABV), size = 2)
```

## Multiple `geom` layers

```{r, echo=TRUE, fig.width=8, fig.height=4}
ggplot(data = nba.data, aes(x = WIN.)) +
  geom_histogram(aes(y = ..density..), binwidth = 0.1, color = "black") +
  geom_density()
```

## Multiple `geom` layers

```{r, echo = TRUE, fig.width=8, fig.height=4}
ggplot(data = nba.data, aes(x = OFFRTG, y = WIN.)) +
  geom_point() + 
  geom_smooth(method = "lm")
```

## Multiple `geom` layers

```{r, echo = TRUE, fig.width=8, fig.height=4}
ggplot(data = nba.data, aes(x = SEASON, y = WIN.)) +
  geom_violin() +
  geom_boxplot(width = 0.2)
```

## The order of `geom` functions is important

```{r, echo = TRUE, fig.width=8, fig.height=4}
ggplot(data = nba.data, aes(x = SEASON, y = WIN.)) +
  geom_boxplot(width = 0.2) +
  geom_violin()
```

## Data transformation {.smaller}

- Sometimes we need to transform the data set to keep variables consistent with the structure of the aesthetics.

- For instance, if we want to compare the mean of winning rate between seasons and regions...

```{r, echo = TRUE}
mean.win <- aggregate(WIN. ~ SEASON + REGION, FUN = mean, data = nba.data)
head(mean.win)
```

Then we can generate a plot to compare the mean of winning rate based on the new data set.

## More options in `geom` function

```{r}

```


## Facet {.smaller}

- Facet function can help you make panel plot very easily

- `facet_wrap` wraps a 1d sequence of panels into 2d.

```{r, echo = TRUE}
p <- ggplot(data = nba.data, aes(x = OFFRTG, y = WIN.)) +
  geom_point() + geom_smooth(method = "lm", se = FALSE)
p + facet_wrap(~SEASON)
```

## Facet {.smaller}

- `facet_grid` forms a matrix of panels defined by row and column faceting variables.

```{r, echo = TRUE}
p <- ggplot(data = nba.data, aes(x = OFFRTG, y = WIN.)) +
  geom_point() + geom_smooth(method = "lm", se = FALSE)
p + facet_grid(REGION ~ SEASON)
```

## Facet {.smaller}

- You can free the scales of x axis and y axis.

```{r, echo = TRUE}
p <- ggplot(data = nba.data, aes(x = OFFRTG, y = WIN.)) +
  geom_point() + geom_smooth(method = "lm", se = FALSE)
p + facet_grid(REGION ~ SEASON, scales = "free")
```

## Facet 

- Be careful when you free your scales...

```{r, fig.width=6, fig.height=4, out.width="50%", fig.align='default'}
p <- ggplot(data = nba.data, aes(x = OFFRTG, y = WIN.)) +
  geom_point() + geom_smooth(method = "lm", se = FALSE)
p + facet_grid(REGION ~ SEASON)
p + facet_grid(REGION ~ SEASON, scales = "free")
```

## Scale

The scale functions control how the plot maps data values to the visual values of an aesthetic, for instance,

`scale_x_continuous`

`scale_y_discrete`

`scale_color_gradient`

`scale_fill_manual`

You can also specify the label of axis or legends in the scale funtion。

## Scale: General Purpose Scales

- `scale_*_continuous`: change the scale for continuous variable

- `scale_*_discrete`: change the scale for discrete variable

- `scale_*_identity`: use values without scaling

- `scale_*_manual`: create your own discrete scale

## Scale: General Purpose Scales

```{r, echo = TRUE}
p <- ggplot(data = nba.data, aes(x = OFFRTG, y = WIN., color = REGION)) + geom_point()
p + scale_x_continuous(name = "offensive rate", limits = c(97, 116)) +
  scale_y_continuous(name = "winning rate", breaks= seq(0, 1, 0.1)) +
  scale_color_manual(name = "region", labels = c("EAST", "WEST"), values = c("blue", "red"))
```

## Scales: X and Y Axis

- General purpose scales work for x and y axis.

```{r, echo = TRUE}
p <- ggplot(data = nba.data, aes(x = OFFRTG, y = WIN., color = REGION)) + geom_point()
p + scale_x_reverse()
```


## Scale: Color & Fill {.smaller}

- General purpose scales also work for color and fill.

- R color cheatsheet: https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/colorPaletteCheatsheet.pdf

- Some scale functions designed to control the scale of color and fill, for instance,

```{r, echo=TRUE, fig.height=3}
p <- ggplot(data = nba.data, aes(x = OFFRTG, y = DEFRTG, color = WIN.)) + geom_point()
p + scale_color_gradient(low = "green", high = "red")
```

## Scale: Size, Shape and Linetype {.smaller}

- General purpose scales also work for size, shape and linetype.

- Reference for shape and linetype code: http://www.cookbook-r.com/Graphs/Shapes_and_line_types/

```{r, echo=TRUE}
p <- ggplot(data = nba.data, aes(x = OFFRTG, y = DEFRTG, shape = REGION)) + geom_point()
p + scale_shape_discrete("Region", solid = FALSE)
```

## Coordinate system {.smaller}

- `coord_*` function control the transformation of the coordinate systems

```{r, fig.show = "hold", out.width = "35%", fig.align='default', echo=TRUE, fig.height=3.5, fig.width=7}
p <- ggplot(data = nba.data, aes(x = OFFRTG, y = WIN.)) +
  geom_point() + geom_smooth(method = "lm")
p + coord_fixed(ratio = 20); p + coord_flip(); p + coord_trans(y = "sqrt"); p + coord_polar()
```


## Theme {.smaller}

- We can change the theme of plot using `theme_*` function

```{r, fig.show = "hold", out.width = "35%", fig.align='default', echo=TRUE, fig.height=3.5, fig.width=7}
p <- ggplot(data = nba.data, aes(x = OFFRTG, y = WIN.)) + 
  geom_point() + geom_smooth(method = "lm")
p + theme_bw(); p + theme_classic(); p + theme_grey(); p + theme_minimal()
```

## Design your own plot

- `theme` function is a powerful way to customize the non-data components of your plots: i.e. titles, labels, fonts, background, gridlines, and legends. See R help for details.

## Arrange and save your plots {.smaller}

- `grid.arrange` from `gridExtra` package can place multiple ggplot on a page

```{r}
p <- ggplot(data = nba.data, aes(x = OFFRTG, y = WIN.)) + 
  geom_point() + geom_smooth(method = "lm")
p1 <- p + theme_bw(); p2 <- p + theme_classic(); p3 <- p + theme_grey(); p4 <- p + theme_minimal()
```

```{r, echo = TRUE, fig.height=3}
grid.arrange(p1, p2, p3, p4, ncol = 2, nrow = 2)
```

- `ggsave` can save the plot to your local drive.

```{r, echo=T, eval=F}
ggsave(p, filename = "", height = , width = , units = )
```

## ggplot2 cheat sheet

- We don't need to memorize all the stuffs. The cheatsheet of ggplot2 gives us a tidy and detailed reference.

https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf

- R help is also a great resource.

## Practice 1

```{r, fig.width=9, fig.height=5.5}
practice1
```

## Practice 2

```{r, fig.width=9, fig.height=5.5}
practice2
```

# Extentions of ggplot2

## Extentions of ggplot2

- GGally: An extention to reduce the complexity of combining geometric objects with transformed data

- ggExtra: A package which can add marginal density plots or histograms to ggplot2 scatterplots.

- ggrepel: A convenient package for `geom_text()`

- ggvis: Interactive plots based on grammar of graphics

- gganimate: A grammar of animated graphics

- more information: http://www.ggplot2-exts.org/gallery/

## GGally

- `ggpairs`: Make a matrix of plots with a given data set.

```{r, echo =  TRUE, fig.width=8, fig.height=4, warning=FALSE}
ggpairs(data = nba.data, 3:7)
```

## GGally

- `ggcorr`: plot a correlation matrix (heatmap) with ggplot2

```{r, echo=TRUE, fig.width=8, fig.height=4}
ggcorr(data = nba.data[, c(3:7)])
```

## ggExtra

- `ggMarginal`: Create a ggplot2 scatterplot with marginal density plots (default) or histograms, or add the marginal plots to an existing scatterplot.

```{r, echo=T}
p <- ggplot(nba.data, aes(x = OFFRTG, y = DEFRTG, color = REGION)) +
  geom_point() + theme_bw() + theme(legend.position = "bottom")
ggMarginal(p, groupColour = TRUE, groupFill = TRUE)
```

## ggrepel {.smaller}

- `geom_text_repel` can solve the problem of overlapping labels when we plot text on the graph.

```{r, echo=T}
ggplot(data = nba.data, aes(x = OFFRTG, y = DEFRTG, size = WIN.)) +
  geom_point(aes(color = REGION), shape = 1) + 
  geom_text_repel(data = subset(nba.data, WIN. >= 0.6 | WIN. <= 0.3), 
                  aes(label = ABV), size = 1.5, box.padding = 0.3) +
  facet_wrap(~SEASON)
```

## ggrepel {.smaller}

- `geom_label_repel` draws a rectangle underneath the text, making it easier to read.

```{r, echo=T}
ggplot(data = nba.data, aes(x = OFFRTG, y = DEFRTG, size = WIN.)) + 
  geom_point(aes(color = REGION), shape = 1) + 
  geom_label_repel(data = subset(nba.data, WIN. >= 0.6 | WIN. <= 0.3), 
                  aes(label = ABV), size = 1.5, box.padding = 0.3) +
  facet_wrap(~SEASON)
```

## ggvis

## gganimate

```{r, echo = T, eval = F}
ggplot(data = nba.data, aes(x = OFFRTG, y = DEFRTG, size = WIN.)) +
  geom_point(aes(color = REGION), shape = 1) + 
  geom_text_repel(aes(label = ABV), size = 1.5, box.padding = 0.3) +
  theme_bw() +
  scale_y_reverse(limits = c(120, 97)) +
  scale_color_manual(values = c("blue3", "red3")) +
  # Here comes the gganimate specific bits
  labs(title = 'SEASON: {closest_state}', x = 'OFFRTG', y = 'DEFRTG') +
  theme(title = element_text(size = 5), 
        text = element_text(size = 2)) +
  transition_states(SEASON,
                    transition_length = 2,
                    state_length = 1)
```

## gganimate

```{r, fig.width = 5.5, fig.height = 2.7}
ggplot(data = nba.data, aes(x = OFFRTG, y = DEFRTG, size = WIN.)) +
  geom_point(aes(color = REGION), shape = 1) + 
  geom_text_repel(aes(label = ABV), size = 1.5, box.padding = 0.3) +
  theme_bw() +
  scale_y_reverse(limits = c(120, 97)) +
  scale_color_manual(values = c("blue3", "red3")) +
  # Here comes the gganimate specific bits
  labs(title = 'SEASON: {closest_state}', x = 'OFFRTG', y = 'DEFRTG') +
  theme(title = element_text(size = 5), 
        text = element_text(size = 5)) +
  transition_states(SEASON,
                    transition_length = 2,
                    state_length = 1)
```

## Reference

# Thanks for listening!
