---
title: "Data Visualization in R"
author: "Yiming Zhang"
date: "10/10/2020"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 7,
                      fig.height = 3.5, 
                      fig.align = "center", warning = FALSE, cache = T)

package.required <- c("ggplot2", "ggrepel", "GGally", "ggExtra", "gganimate", "gridExtra", "gifski", "datasauRus", "dplyr")

for(i in 1:length(package.required)){
  if(!package.required[i] %in% rownames(installed.packages())){
    install.packages(package.required[i])
  }
}

## Load packages
library(ggplot2)
library(ggrepel)
library(ggExtra)
library(GGally)
library(gganimate)
library(gridExtra)
library(gifski)
library(datasauRus)
library(dplyr)

## Read data
nba.data <- read.csv("NBA_Team_Data.csv")
nba.data <- nba.data[, c(2, 3, 7, 9, 30, 31, 42, 44, 45)]

practice1 <- ggplot(data = nba.data, aes(x = OFFRTG, y = DEFRTG, size = WIN.)) +
  geom_point(aes(color = REGION), shape = 1) + 
  geom_text(data = subset(nba.data, WIN. > 0.65),
            aes(label = ABV), size = 1.5) +
  geom_text_repel(data = subset(nba.data, WIN. < 0.3),
            aes(label = ABV), size = 1.5, 
            min.segment.length = 0, box.padding = 0.3) +
  facet_wrap(~SEASON) +
  theme_bw() +
  scale_x_continuous("Offensive Rate") +
  scale_y_reverse("Defensive Rate", limits = c(118, 95)) +
  scale_color_manual("Region" ,values = c("blue3", "red3")) +
  scale_size_continuous("Winning Rate", breaks = c(0.2, 0.4, 0.6)) +
  theme(legend.position = "bottom",
    panel.grid.minor = element_blank())

dinodata <- subset(datasaurus_dozen, dataset != "away")
summary.dino <- dinodata %>%
  group_by(dataset) %>%
  summarise(meanx = round(mean(x),3), 
            meany = round(mean(y),3),
            sdx = round(sd(x),3), 
            sdy = round(sd(y),3),
            corxy = round(cor(x,y),3))
```

## Outline

- General introduction to data visualization

- Introduction to ggplot2

 + Grammar of graphics
 
 + A case study using NBA data

- Useful packages and extensions for ggplot2

## WHEN and WHY to visualize data

- Exploratory data analysis

  + Explore pattern, trend, and distribution of one variable
  
  + Explore association between variables
  
- Statistical analysis

  + Diagnostic plots for linear regression

- Report your results and communicate with non-statisticians

  + A more clear way of presenting findings

  + Attract your audiences

- For fun...

## An example for fun

```{r, fig.width=9, fig.height = 5.5}
ggplot(dinodata, aes(x=x, y=y))+
  geom_point(alpha=0.5)+
  theme(legend.position = "none")+
  facet_wrap(~dataset, ncol=3)
```

## An example for fun

```{r}
summary.dino
```

A statistical plot contains much more information than a table of summary statistics!

## What to plot

- One variable: Histogram, Bar chart, Density plot...

- Two variables: Scatter plot, Box plot, Violin Plot...

- Multiple variables: Heatmap...

- Checking normality: qqplot...

- Think of your data and variables carefully, and choose the most appropriate statistical plot.

## Introduction to the data set {.smaller}

- All data are from http://stats.nba.com

```{r}
head(nba.data)
```

- WIN.: Winning rate, which is the percentage of games played that a team has won.

- PTS: The number of points scored.

- OFFRTG: Offensive Rating, which measures a team's points scored per 100 possessions.

- DEFRTG: Defensive Rating, which is the number of points allowed per 100 possessions by a team.

- PACE: Pace, which is the number of possessions per 48 minutes for a team.

- REGION: East/West.

- ABV: The abbreviation of a team.

## A "good" plot can deliver a lot of information

```{r,  fig.width=9, fig.height = 5.5}
practice1  
```

## But "bad" plots may...

- be hard to read if labels and legends are not clear

- confuse people if it is not well-designed

- deliver misleading information (sometimes in purpose)

## Visualization tools in R - A histogram example 

The histograms of winning rate in different regular NBA seasons and regions generated by `ggplot2` and `graphics` packages:

```{r, echo=FALSE, fig.show = "hold", out.width = "50%", fig.align='default', fig.height = 5.5}

sub.dt <- subset(nba.data, SEASON %in% c("2017-2018", "2018-2019"))
sub.dt$SEASON <- factor(as.character(sub.dt$SEASON))

ggplot(data = sub.dt, aes(x = WIN.)) + 
  geom_histogram(binwidth = 0.1, color = "black") + facet_grid(REGION ~ SEASON)

par(mfrow = c(2, 2), mar = c(2, 2, 3, 1))
for(i in levels(sub.dt$REGION)){
  for(j in levels(sub.dt$SEASON)){
    subdata <- subset(sub.dt, REGION == i & SEASON == j)
    hist(subdata$WIN., breaks = seq(0.05, 0.95, 0.1),
         xlim = c(0, 1),
         main = paste(i, j, sep = " ,"))
  }
}
```

## Comparing codes for the same plot

Code in `ggplot2`:

```{r, echo = TRUE, eval = FALSE}
ggplot(data = sub.dt, aes(x = WIN.)) + 
  geom_histogram(binwidth = 0.1, color = "black") + facet_grid(REGION ~ SEASON)
```

Code in `graphics` package

```{r, echo = TRUE, eval = FALSE}
par(mfrow = c(2, 2), mar = c(2, 2, 3, 1))
for(i in levels(sub.dt$REGION)){
  for(j in levels(sub.dt$SEASON)){
    subdata <- subset(sub.dt, REGION == i & SEASON == j)
    hist(sub.dt$WIN., breaks = seq(0, 1, 0.1),
         main = paste(i, j, sep = " ,"))
  }
}
```

## Grammar of Graphics {.smaller}

- Idea: graph is a combination of independent building blocks.

- **Data** that you want to visualize and a set of **aesthetic** mappings describing how variables in the data are mapped to aesthetic attributes.

- **Layers** made up of geometric elements and statistical transformation. Geometric objects, **geoms** for short, such as points, lines, polygons, etc. Statistical transformations, **stats** for short, summarize data in many useful ways.

- The **scales** map values in the data space to values in an aesthetic space, whether it be color, or size, or shape.

- A **coordinate system**, coord for short, describes how data coordinates are mapped to the plane of the graphic.

- A **facet** describes how to break up the data into subsets and how to display those subsets as small multiples. 

- A **theme** which controls the finer points of display, like the font size and background color. 

## The start of plotting a graph {.smaller}

- `ggplot()` is always the first line of your code.

- We can specify the **data** set and the **aesthetics** mapping variables in the `ggplot()`.

```{r, echo=TRUE}
p <- ggplot(data = nba.data, aes(x = OFFRTG, y = WIN.))
p
```

## Aesthetics

- Map the variables in the data to the components in the plot

- x: x axis

- y: y axis

- color: color of the boundary of a symbol

- fill: color of the inside of a symbol

- shape: shape of points, solid point, circle, triangle...

- size: size of points

- linetype: type of lines, solid line, dashed line...

- ...

## Geometries {.smaller}

- Geometries are the actual graphical elements displayed in a plot. They can visualize the mapping variables (specified in `aes()`) from the data.

- We use `+` to connect multiple geometry functions

```{r, echo = TRUE}
p + geom_point()
```

## Geometries

- We can also specify `data` and `aes` in `geom` function. They don't have to be the same as those in `ggplot()`.

```{r, echo = TRUE}
ggplot() + geom_point(data = nba.data, aes(x = DEFRTG, y = WIN.))
```

## `geom` function {.smaller}

- One continuous variable

```{r, fig.show = "hold", out.width = "50%", fig.align='default', echo=TRUE}
p <- ggplot(data = nba.data, aes(x = WIN.))
p + geom_histogram(binwidth = 0.1)
p + geom_density()
```

## `geom` function {.smaller}

- Continuous X, continuous Y

```{r, fig.show = "hold", out.width = "35%", fig.align='default', echo=TRUE, fig.height=3.5, fig.width=7}
p <- ggplot(data = nba.data, aes(x = OFFRTG, y = WIN.))
p + geom_point(); p + geom_line(); p + geom_density_2d(); p + geom_smooth(formula = y ~ x, method = "lm")
```

## `geom` function {.smaller}

- Discrete X, continuous Y

```{r, fig.show = "hold", out.width = "50%", fig.align='default', echo=TRUE}
p <- ggplot(data = nba.data, aes(x = SEASON, y = WIN.))
p + geom_boxplot()
p + geom_violin()
```

## Multiple `geom` layers

```{r, echo = TRUE, fig.width=8, fig.height=4}
ggplot(data = nba.data, aes(x = WIN.)) +
  geom_histogram(aes(y = ..density..), binwidth = 0.1, color = "black") +
  geom_density()
```

## Multiple `geom` layers

```{r, echo = TRUE, fig.width=8, fig.height=4}
ggplot(data = nba.data, aes(x = OFFRTG, y = WIN.)) +
  geom_point() + 
  geom_smooth(formula = y ~ x, method = "lm")
```

## Multiple `geom` layers

```{r, echo = TRUE, fig.width=8, fig.height=4}
ggplot(data = nba.data, aes(x = SEASON, y = WIN.)) +
  geom_violin() +
  geom_boxplot(width = 0.2)
```

## The order of `geom` functions is important

```{r, echo = TRUE, fig.width=8, fig.height=4}
ggplot(data = nba.data, aes(x = SEASON, y = WIN.)) +
  geom_boxplot(width = 0.2) +
  geom_violin()
```

## Facet {.smaller}

- Facet function can help you make panel plot very easily

- `facet_wrap` wraps a 1d sequence of panels into 2d.

```{r, echo = TRUE}
p <- ggplot(data = nba.data, aes(x = OFFRTG, y = WIN.)) +
  geom_point() + geom_smooth(formula = y ~ x, method = "lm", se = FALSE)
p + facet_wrap(~SEASON)
```

## Facet {.smaller}

- `facet_grid` forms a matrix of panels defined by row and column faceting variables.

```{r, echo = TRUE}
p <- ggplot(data = nba.data, aes(x = OFFRTG, y = WIN.)) +
  geom_point() + geom_smooth(formula = y ~ x, method = "lm", se = FALSE)
p + facet_grid(REGION ~ SEASON)
```

## Scale {.smaller}

- The scale functions control how the plot maps data values to the visual values of an aesthetic, for instance,

  + `scale_x_continuous`

  + `scale_y_discrete`

  + `scale_color_gradient`

  + `scale_fill_manual`

- The format of scale functions is always `scale_element1_element2`. The first element represents the aesthetics, and the second element represents the characteristics of variables.

- You can also specify the label of axis or legends in the scale function.

- R color cheat sheet: https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/colorPaletteCheatsheet.pdf

## Scale

```{r, echo = TRUE, fig.height=3}
p <- ggplot(data = nba.data) + 
  geom_point(aes(x = OFFRTG, y = DEFRTG, color = WIN., shape = REGION))
p + scale_x_continuous(name = "offensive rate", limits = c(97, 116)) +
  scale_y_reverse(name = "defensive rate") +
  scale_color_gradient(name = "winning rate", low = "green", high = "red") +
  scale_shape_discrete(name = "region", labels = c("EAST", "WEST"))
```

## Design your own plot

- `coord_*` function control the transformation of the coordinate systems, such as `coord_trans(y = "sqrt")`.

- We can change the theme of plot using `theme_*` function

- `labs` function can set the title, subtitle and caption of your plot.

- `theme` function is a powerful way to customize the non-data components of your plots: i.e. titles, labels, fonts, background, grid lines, and legends. See R help for details.

- `ggsave` can save the plot to your local drive.

## ggplot2 online documents

- We don't need to memorize all the stuffs. The cheat sheet of ggplot2 gives us a tidy and detailed reference.

https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf

- R help is also a great resource.

- [Top 50 ggplot2 Visualizations](http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html)

- Wickham, H. (2016). ggplot2: elegant graphics for data analysis. Springer.

## A case study

```{r, fig.width=9, fig.height=5.5}
practice1  
```

## A case study {.smaller}

```{r, echo = T, eval = F}
ggplot(data = nba.data, aes(x = OFFRTG, y = DEFRTG, size = WIN.)) +
  geom_point(aes(color = REGION), shape = 1) + 
  geom_text(data = subset(nba.data, WIN. > 0.65),
            aes(label = ABV), size = 1.5) +
  geom_text_repel(data = subset(nba.data, WIN. < 0.3),
            aes(label = ABV), size = 1.5, 
            min.segment.length = 0, box.padding = 0.3) +
  facet_wrap(~SEASON) +
  theme_bw() +
  scale_x_continuous("Offensive Rate") +
  scale_y_reverse("Defensive Rate", limits = c(118, 95)) +
  scale_color_manual("Region" ,values = c("blue3", "red3")) +
  scale_size_continuous("Winning Rate", breaks = c(0.2, 0.4, 0.6)) +
  theme(legend.position = "bottom",
    panel.grid.minor = element_blank())
```


## Useful packages or extensions for ggplot2

- gridExtra: A package can help you arrange multiple plots on a page

- GGally: An extension to reduce the complexity of combining geometric objects with transformed data

- ggExtra: A package which can add marginal density plots or histograms to ggplot2 scatter plots.

- ggrepel: A convenient package for `geom_text()`

- gganimate: A grammar of animated graphics

- more information: https://exts.ggplot2.tidyverse.org/gallery/

## GGally

- `ggpairs`: Make a matrix of plots with a given data set.

- `ggcorr`: plot a correlation matrix (heatmap) with ggplot2

```{r, echo =  TRUE, warning=FALSE, fig.show = "hold", out.width = "50%", fig.align='default'}
ggpairs(data = nba.data, 3:7)
ggcorr(data = nba.data[, 3:7])
```

## ggExtra

- `ggMarginal`: Create a ggplot2 scatter plot with marginal density plots (default) or histograms, or add the marginal plots to an existing scatter plot.

```{r, echo=T}
p <- ggplot(nba.data, aes(x = OFFRTG, y = DEFRTG, color = REGION)) +
  geom_point() + theme_bw() + theme(legend.position = "bottom")
ggMarginal(p, groupColour = TRUE, groupFill = TRUE)
```

## gganimate

```{r, fig.width = 5.5, fig.height = 2.7}
ggplot(data = nba.data, aes(x = OFFRTG, y = DEFRTG, size = WIN.)) +
  geom_point(aes(color = REGION), shape = 1) + 
  geom_text_repel(aes(label = ABV), size = 1.5, box.padding = 0.3) +
  theme_bw() +
  scale_y_reverse(limits = c(120, 97)) +
  scale_color_manual(values = c("blue3", "red3")) +
  # Here comes the gganimate specific bits
  labs(title = 'SEASON: {closest_state}', x = 'OFFRTG', y = 'DEFRTG') +
  theme(title = element_text(size = 5), 
        text = element_text(size = 5)) +
  transition_states(SEASON,
                    transition_length = 2,
                    state_length = 1)
```

## gganimate

```{r, echo = T, eval = F}
ggplot(data = nba.data, aes(x = OFFRTG, y = DEFRTG, size = WIN.)) +
  geom_point(aes(color = REGION), shape = 1) + 
  geom_text_repel(aes(label = ABV), size = 1.5, box.padding = 0.3) +
  theme_bw() +
  scale_y_reverse(limits = c(120, 97)) +
  scale_color_manual(values = c("blue3", "red3")) +
  # Here comes the gganimate specific bits
  labs(title = 'SEASON: {closest_state}', x = 'OFFRTG', y = 'DEFRTG') +
  theme(title = element_text(size = 5), 
        text = element_text(size = 2)) +
  transition_states(SEASON,
                    transition_length = 2,
                    state_length = 1)
```

# Thanks for listening!
